# Creating a ggplot

The best place to continue learning about ggplot is the **ggplot2 book**: [ggplot2: Elegant graphics for data analysis](https://ggplot2-book.org/).

## Define a Plot Object

With **ggplot2**, you begin a plot with the function `ggplot()`, defining a plot object that you then add layers to.

The first argument of `ggplot()` is the **dataset** to use in the graph and so `ggplot(data = penguins)` creates an empty graph that is primed to display the penguins data, but since we haven’t told it how to visualize it yet, for now it’s empty (it's like an empty canvas you’ll paint the remaining layers of your plot onto).

*Example:*

```{r echo = FALSE, message = FALSE}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
```

```{r}
ggplot(data = penguins)
```

## Map the Data

Next, we need to tell `ggplot()` how the information from our data will be visually represented. The `mapping` argument of the `ggplot()` function defines how variables in your dataset are mapped to visual properties (aesthetics) of your plot.

The `mapping` argument is always defined in the `aes()` function, and the `x` and `y`arguments of `aes()` specify which variables to map to the x and y axes.

*Example:*

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
)
```

Our empty canvas now has more structure – it’s clear where flipper lengths will be displayed (on the x-axis) and where body masses will be displayed (on the y-axis). But the penguins themselves are not yet on the plot. This is because we have not yet determine how to represent the observations from our dataframe on our plot.

## Represent the Data

To do so, we need to define a `geom`: the geometrical object that a plot uses to represent data. These geometric objects are made available in ggplot2 with functions that start with `geom_`.

For example, bar charts use `geom_bar()`, line charts use `geom_line()`, boxplots use `geom_boxplot()`, scatterplots use `geom_point()`, and so on.

*Example:*

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```

**Note:** One common problem when creating ggplot2 graphics is to put the `+` in the wrong place: it has to come at the end of the line, not the start.

You can **rewrite the previous plot** more concisely:

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()
```

Or using the pipe, `|>`:

```{r}
penguins |> 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()
```

Now we have something that looks like what we might think of as a “scatterplot”.

You can also set the visual **properties of your geom** manually as an argument of your `geom` function (outside of `aes()`) instead of relying on a variable mapping to determine the appearance. For example, we can make all of the points in our plot blue:

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(color = "blue")
```

### Help with geoms

The best place to get a comprehensive **overview of all of the geoms** ggplot2 offers, as well as all functions in the package, is this page: <https://ggplot2.tidyverse.org/reference>.

To learn more about any single geom, use the help (e.g., `?geom_smooth`).

See also <https://exts.ggplot2.tidyverse.org/gallery/> for a sampling of community made geoms.

## Add Aesthetics and Layers

You will we need to modify the aesthetic mapping, inside of `aes()`.

*Example:*

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point()
```

Once you **map an aesthetic**, ggplot2 takes care of the rest. It selects a reasonable scale to use with the aesthetic, and it constructs a **legend** that explains the mapping between levels and values. For `x` and `y` aesthetics, ggplot2 does not create a legend, but it creates an **axis line** with tick marks and a label. The axis line provides the same information as a legend; it explains the mapping between locations and values.

When a **categorical variable** is mapped to an aesthetic, ggplot2 will automatically assign a unique value of the aesthetic (here a unique color) to each unique level of the variable (each of the three species), a process known as scaling. ggplot2 will also add a **legend** that explains which values correspond to which levels.

Now let’s add **one more layer**: a **smooth curve** displaying the relationship between body mass and flipper length. Since this is a new geometric object representing our data, we will add **a new geom** as a layer on top of our point geom: `geom_smooth()`. And we will specify that we want to draw the line of best fit based on a linear model with `method = "lm"`.

*Example:*

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point() +
  geom_smooth(method = "lm")
```

When aesthetic mappings are defined in `ggplot()`, at the **global level**, they’re passed down to each of the subsequent geom layers of the plot. However, each `geom` function in ggplot2 can also take a mapping argument, which allows for aesthetic mappings at the **local level** that are added to those inherited from the global level. In other words, if you place mappings in a `geom` function, ggplot2 will treat them as local mappings for the layer. It will use these mappings to extend or overwrite the global mappings for that layer only. This makes it possible to display different aesthetics in different layers.

*Example:*

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species)) +
  geom_smooth(method = "lm")
```

Note that since we want points to be colored based on species but don’t want the lines to be separated out for them, we should specify `color = species` for `geom_point()` only.

You can specify different **data for each layer**. Here, we use red points as well as open circles to highlight two-seater cars. The local data argument in `geom_point()` overrides the global data argument in `ggplot()` for that layer only.

*Example:*

``` {r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_point(
    data = mpg |> filter(class == "2seater"), 
    color = "red"
  ) +
  geom_point(
    data = mpg |> filter(class == "2seater"), 
    shape = "circle open", size = 3, color = "red"
  )
```

It’s generally not a good idea to represent information using only colors on a plot, as people perceive colors differently due to color blindness or other color vision differences. Therefore, in addition to color, we can also map species to the **shape aesthetic**.

*Example:*

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = "lm")
```

### Help with aes

You can learn more about all possible **aesthetic mappings** in the aesthetic specifications vignette at <https://ggplot2.tidyverse.org/articles/ggplot2-specs.html>. Remember that the specific aesthetics you can use for a plot depend on the geom you use to represent the data.

## Facets

**Facets** splits a plot into subplots that each display one subset of the data based on a **categorical variable**.

*Example:*

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_wrap(~cyl)
```

To facet your plot with the combination of two variables, switch from `facet_wrap()` to `facet_grid()`. The first argument of `facet_grid()` is also a formula, but now it’s a double sided formula: `rows ~ cols`.

*Example:*

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv ~ cyl)
```

By default each of the facets share the same **scale** and **range** for `x` and `y` axes. This is useful when you want to compare data across facets but it can be limiting when you want to visualize the relationship within each facet better. Setting the scales argument in a faceting function to `free_x` will allow for different scales of x-axis across columns, `free_y` will allow for different scales on y-axis across rows, and `free` will allow both.

*Example:*

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv ~ cyl, scales = "free")
```
  
## Improve Labels

And finally, we can improve the **labels** of our plot using the `labs()` function in a new layer:

-   `title`, adds a title.
-   `subtitle`, adds a subtitle to the plot.
-   `x`, is the x-axis label.
-   `y`, is the y-axis label.
-   `color` and `shape` define the label for the legend.

In addition, we can improve the color palette to be **colorblind safe** with the `scale_color_colorblind()` function from the **ggthemes package**.

*Example:*

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  ) +
  scale_color_colorblind()
```

## Save Your Plot

Once you’ve made a plot, you might want to get it out of R by saving it as an **image** that you can use elsewhere. That’s the job of `ggsave()`, which will save the plot most recently created to disk:

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
ggsave(filename = "penguin-plot.png")
```

This will save your plot to your **working directory**.
