# Reading Data

R provides a number of ways to read data from a file, the **most flexible** of which is the `scan()` function. We use scan to read a vector of values from a file. It has the form:

*Example:*

``` r
scan(file = "", what = 0, n = -1, sep = "", skip = 0, quiet = FALSE)
```

Remember that `scan()` returns a **vector**.

## Tabular Data

It is common for data to be arranged in **tables**, with columns corresponding to variables and rows corresponding to separate observations. These **dataframes** are usually read into R using the function `read.table()`, which has the form:

``` r
read.table(file, header = FALSE, sep = "") 
# read.table() returns a dataframe
```

`read_table()` reads a common variation of fixed-width files where columns are separated by white space.

There are two commonly used **variants** of `read.table()`:

### `read.csv()`

`read.csv()` is for comma-separated data and is equivalent to `read.table(file, header = TRUE, sep = ",")`.

Sometimes, it could happen that the file extension is **.csv**, but the data is **not comma separated**. In that case, we can still use the `read.csv()` function, but in this case we have to specify the separator:

*Example:*

``` r
read.csv("iris_semicolon.csv", stringsAsFactors = FALSE, sep=";")
anscombe_tab_2 <- read.table("anscombe.txt", header=TRUE)
```

Usually, `read_csv()` uses the **first line** of the data for the column names, which is a very common convention. But it’s not uncommon for a few lines of metadata to be included at the top of the file. You can use `skip = n` to skip the first n lines or use `comment = "#"` to drop all lines that start with (e.g.) `#`.

*Example:*

``` r
read_csv("data/students.csv,
  skip = 2,
  comment = "#"
)
```

In other cases, the data might **not have column names**. You can use `col_names = FALSE` to tell `read_csv()` not to treat the first row as headings and instead label them sequentially from $X_1$ to $X_n$. Alternatively, you can pass `col_names` a character vector which will be used as the column names (e.g. `col_names = c("x", "y", "z")`):

Note that by default `read_csv()` only recognizes empty strings ("") in as NAs, however you surely want it to also recognize the **character string "N/A"**.

*Example:*

``` r
students <- read_csv("data/students.csv", na = c("N/A", ""))
```

If a **.csv file** contains both numeric and character variables, and we use `read.csv()`, the character variables get automatically converted to the **factor** type. We can prevent character variables from this automatic conversion to factor, by specifying `stringsAsFactors=FALSE` within the `read.csv()` function.

*Example:*

``` r
read.csv("iris.csv", stringsAsFactors=F)
```

**readr** provides a total of nine column types for you to use: `col_logical()`, `col_logical()`, `col_factor()`... Find more information [here](https://readr.tidyverse.org/https://readr.tidyverse.org/).

### `read.delim()`

`read_delim()` reads in files with any delimiter, attempting to automatically guess the delimiter if you don’t specify it. For example, `read.delim()`is for tab-delimited data and is equivalent to `read.table(file, header = TRUE, sep = "\t")`.

*Example:*

``` r
## skips the first 2 lines
anscombe <- read.csv("CSVanscombe.csv", skip=2) 
```

### Other file types

Once you’ve mastered `read_csv()`, using **readr’s** other functions is straightforward; it’s just a matter of knowing which function to reach for:

-   `read_csv2()`: reads semicolon-separated files. These use `;` instead of `,` to separate fields and are common in countries that use `,` as the decimal marker.

-   `read_tsv()`: reads tab-delimited files.

-   `read_fwf()`: reads fixed-width files. You can specify fields by their widths with `fwf_widths()` or by their positions with `fwf_positions()`.

-   `read_log()`: reads Apache-style log files.

## Reading Multiple Files

Sometimes your data is split across **multiple files** instead of being contained in a single file. For example, you might have sales data for multiple months, with each month’s data in a separate file: `01-sales.csv` for January, `02-sales.csv` for February, and `03-sales.csv` for March.

With `read_csv()`you can read these data in at once and stack them on top of each other in a single data frame.

*Example:*

``` r
sales_files <- c("data/01-sales.csv", "data/02-sales.csv", "data/03-sales.csv")
read_csv(sales_files, id = "file")
```

The `id` argument adds a new column called file to the resulting data frame that identifies the file the data come from. This is especially helpful in circumstances where the files you’re reading in do not have an identifying column that can help you trace the observations back to their original sources.

An alternative to read multiple data made **available online** is this:

*Example:*

``` r
sales_files <- c(
  "https://pos.it/r4ds-01-sales",
  "https://pos.it/r4ds-02-sales",
  "https://pos.it/r4ds-03-sales"
)
read_csv(sales_files, id = "file") 
```

## Reading Line by Line

Text files can be **read line by line** using the `readLines()` function:

*Example:*

``` r
con <- gzfile("words.gz") 
x <- readLines(con, 10)
```

This approach is useful because it allows you to read from a file **without having to uncompress the file first**, which would be a waste of space and time.

To read in lines of **webpages** it can be useful the `readLines()` function:

*Example:*

``` r
con <- url("http://www.jhsph.edu", "r") ## Read the web page 
x <- readLines(con) ## Print out the first few lines head(x)
```

## xls Format

If the dataset is stored in the ***.xls or*** **.xlsx format** you can use `read.xlsx2`:

*Example:*

``` r
library(xlsx) 
anscombe_xlsx <- read.xlsx2("xlsxanscombe.xlsx", sheetIndex=1)
```

## RData Format

If you need to store more than one dataset in a single file we can use the **\*.RData format**:

*Example:*

``` r
## example to load multiple datasets, and a vector of R objects from a single *.RData
load("robjects.RData")
## to check if objects have been loaded correctly
objects()
```

## Import Stata/SPSS File

To import a **Stata/SPSS file** into R:

-   First you need to call the foreign library:

    ``` r
    install.packages("foreign")
    ```

-   Then use `read.data()`/`read.spss()`:

    ``` r
    data <- read.spss(file="data.spss", to.data.frame=TRUE)
    ```

-   The output will always be a data frame:

    ``` r
    write.foreign(data, "mydata.txt", "mydata.sps", package="SPSS")
    ```

## JSON File

To read a **JSON file**:

``` r
install.packages("rjson") 
data <- fromJSON(file="data.json")
data2 <- as.data.frame(data)
```

## `view()`

To view the data variable, you can use the `view()`.

## Manual Data Entry

Sometimes you’ll need to assemble a **tibble “by hand”** doing a little data entry in your R script. 

There are **two useful functions** to help you do this which differ in whether you layout the tibble by columns or by rows. 

-   `tibble()` works by column:

    *Example:*
    
    ``` {r}
    tidyr::tibble(
      x = c(1, 2, 5), 
      y = c("h", "m", "g"),
      z = c(0.08, 0.83, 0.60)
    )   
    ```

-   `tribble()` (transposed tibble) lets you lay out your data row by row: column headings start with `~` and entries are separated by commas. This makes it possible to lay out small amounts of data in an easy to read form:

    *Example:*
  
    ``` {r}
    tidyr::tribble(
      ~x, ~y, ~z,
      1, "h", 0.08,
      2, "m", 0.83,
      5, "g", 0.60
    )
    ```
